---
title: "Cooties Survival"
author: "Alex Pan"
date: "April 19, 2018"
output: 
    html_document:
        keep_md: true
        fig_height: 5
        fig_width: 7
        theme: spacelab
        toc: yes
        toc_depth: 4
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, results = 'hold')
options(width = 200)
# setwd('R/R_Instructional/Rudimentary_Survival_Analysis')
```

```{r, results = FALSE}
library(tidyverse)    # Functions to help with data manipulation
library(lubridate)    # Functions to help handle dates
library(survival)    # Statistical package for analyzing survival data
library(survminer)
```

## Introduction

Cooties is a life-threatening pediatric illness. It primarily affects children and adolescents below the age of 15. The 5-year survival of cooties is about 25% if it is allowed to take its natural course. A new drug, COOTICURE, is being tested in clinical trials for efficacy in delaying death. In early clinical trials the therapy is proved quite effective, but only certain people benefited.

We will explore these data (in a completely statistically unsound way) using R, and see what we discover.

***

## Working in R


### Loading in the Data

Let's start by loading in the data from Excel.

The `read_csv()` command will read a '.csv' file (comma separated values) and automatically convert it to an R dataframe. A dataframe in R is like an Excel spreadsheet, but it tends to follow a very specific format.


In an R data frame:

* Each *observation* (e.g. study subject) corresponds to a *row*.

* Each *variable* (data about the subjects) corresponds to a *column*.

* The intersection between a row and a column is a *cell*.

* The value in a cell (eg., `id = 1` is referred to as an **element** or an **entry**.


```{r}
# When we begin a line with a '#', it means that we are writing a comment.
# Comments are only to communicate to other humans; they are ignored by the program!
cooties <- read_csv('cooties_data.csv')
```

We've stored the data in a dataframe called `cooties`

We can look at the first 10 data entries by *subsetting* or *slicing* the dataframe:

* The [ ] notation denotes that we are subsetting the dataframe `cooties`

* The '1:10' is denotes: "Take rows 1 to 10"

* The empty space after the comma denotes: "Take all of the columns" 
```{r}
cooties[1:10, ]
```

### Data Dictionary

Variable names are limited, and sometimes its not clear what exactly they mean. To work with data, you often need a *data dictionary*. Here is one for you:

`id` Unique patient identification number. No two patients have the same `id`.

`sex` 0 is female, 1 is male

`date of birth` The patient's birthdate, provided (annoyingly) as a text field

`treatment_arm` 0 is placebo, and 1 is COOTICURE

`trial_entrance` The patient's start date in the trial

`death` TRUE if the patient died during the trial

`trial_exit` The patient's last follow-up in the trial

`trial_exit`. 1 If the patient died, and 0 otherwise.


***

### A Quick Introduction to the R tidyverse

The tidyverse is a collection of R packages that are designed for working with data. The tidyverse makes manipulating data easier because:

* Every function is designed to do one thing

* Every function has a consistent format, where the first argument is always a data frame and the output is always a data frame.

* The functions can be chained, rather than nested, which makes it more similar to English, and therefore easier to read. This will be explained later.

Some key tools that we will use from the tidyverse are below. You can learn more in-depth about the tidyverse <a href="https://www.datacamp.com/community/tutorials/tidyverse-tutorial-r">here</a>

***

#### `filter`

Keeps rows that meet certain conditions, and filters out the ones that don't. For example, we can filter the `cooties` data to only include Males in the Placebo treatment arm.

Note that the `cooties` dataframe itself doesn't actually change unless we re-assign the value using the `<-` operator.
```{r}
filter(cooties, sex == 1, treatment_arm == 0)
```

***

#### `select`

Allows you to select the columns that you provide as arguments, or remove columns if you put a `-` in front. For example, if we want to make a new data frame with only `id`, `sex` and `date_of_birth` we can type:
```{r}
select(cooties, id, sex, date_of_birth)
```

But typing this produces the exact same results:
```{r}
select(cooties, -treatment_arm, -trial_entrance, -death, -trial_exit)
```

***

#### `mutate`

Allows you to create new variables from old variables, effecively "mutating" the data frame. For example, we can make a new variable `years_in_trial` to describe how long the subject was followed. We do that by subtracting the end date (`trial_exit`) from the start date (`trial_entrance`).

Because we actually want to keep this change, we will use the assignment operator `<-` to overwrite `cooties`
```{r}
cooties <- mutate(cooties, years_in_trial = as.numeric(trial_exit - trial_entrance) / 365.25)
cooties
```

You also have the option to overwrite variables by replacing the name, as in:
```{r}
mutate(cooties, sex = ifelse(sex == 1, 'Male', 'Female'))


# Note:
# The `ifelse()` function checks each observation to see if it meets a
# condition. Here the condition is that `sex == 1`
# IF it is True: It will return the value 'Male'. 
# ELSE (i.e., it is False): It will return the value 'Female'
```

***

#### `separate`

Allows you to separate a single column into multiple columns based on a "separator"

For example, we can split up the `date_of_birth` column into three different columns of `dob_month`, `dob_day`, and `dob_year`.

```{r}
# If your function is very long, it is easier to read if you separate each
# argument onto a separate line
separate(cooties, 
         col = date_of_birth, 
         into = c('dob_month', 'dob_day', 'dob_year'),
         sep = ' ',    # The 'separator' is a space (as in spacebar)
         remove = TRUE)    # remove = TRUE will remove the original variable

# Note:
# `c()` means 'concatenate'; it is how you attach multiple objects together as
# a single object. Try: 1 + c(1, 2, 3)
```

This is a rather contrived example, but the `separate` function is often very useful when there are unwanted symbols (e.g., lab results reported like: 3.0ug/mL) or for cleaning up dates that are in non-standard formats.


***

#### `%>%`

This weird-looking thing is the 'pipe' or 'chain' operator. As mentioned above, the tidyverse supports chaining, which makes code much much easier to read. Chaining takes the *output* of the last function and uses it as the *input* for the next function (specifically, it is used as the first argument of the next function).

Compare these two code snippets that perform identical steps. Which is easier to follow? 

With Chaining:
```{r}
# Note the manipulations I am doing in these snippets are useless; it is just to illustrate the point

cooties %>% 
    filter(death == TRUE) %>%
    select(-sex) %>%
    mutate(treatment_arm = ifelse(treatment_arm == 1, 'Cooticure', 'Placebo'))
```

Without chaining (i.e., "nested")
```{r}
mutate(
    select(
        filter(cooties, death == TRUE), 
        -sex), 
    treatment_arm = ifelse(treatment_arm == 1, 'Cooticure', 'Placebo'))
```


***

### Exploring Data with R

R is also a very useful for data analysis because of how quickly you can summarize data. You can get descriptive statistics and visualizations very easily to help understand your data.

Let's look at a few examples:

***

`dim` stands for 'dimensions'. It let's us summarize the dimensions of out data frame so we now how big our dataset is. It outputs the number of rows (how many observations you have) and the number of columns (how many variables you have).
```{r, results = 'hold'}
dim(cooties)
```

***

`mean` calculates the mean (as one might expect). `median` and `mode` are also avilable.

`$` (*dollar*) an operator that selects a variable from a dataframe, i.e., *From the data frame `cooties`, select the variable `sex`*. You can think of it as a short-hand way of writing `select(cooties, sex)` (though it is not *exactly* the same)

```{r}
mean(cooties$sex)
# Since sex is a binary variable (1 being male, 0 being female), we can quickly
# calculate the proportion of males by taking the mean. Here it is about 50%.
```

***

`summary` outputs a whole bunch of useful descriptive statistics that help describe the distribution of the variable.
```{r}
summary(cooties$years_in_trial)
```

***

`hist` plots a histogram. A histogram allows you to see the distribution of your data---it plots how frequent each of the values are. This is a painstaking process in Excel, but is done in one line in R.
```{r, fig.height = 4, fig.width = 4}
hist(cooties$years_in_trial)
```


***

`boxplot` plots a boxplot, which is another way that you can visualize the distribution of your data. Boxplots are particularly useful for visually comparing two distributions. For example, we can look at the differences between males and females in terms of their length on the trial.

Note: The `y ~ x` notation is called a "formula". It is not necessary to fully understand this now, but here is how it works: R interprets `years_in_trial` as the response variable (i.e., the *dependent variable*), and `cooties$sex` as a predictor (i.e., an *independent variable*). Formulas are used a lot in statistical modeling functions to communicate what variable we are trying to estimate in the model and what variables we want to use to help estimate it. You can learn more <a href="http://www.dummies.com/programming/r/how-to-use-the-formula-interface-in-r/">here</a>.

```{r, fig.height = 4, fig.width = 4}
boxplot(cooties$years_in_trial ~ cooties$sex)
```


These quick plots get the job done, but it's not pleasing to the eye. In later sessions, we will go over how to create more professional-looking plots.


### Survival Analysis
 
Create a model for `death ~ treatment_arm`
 
```{r}
survival <- survfit(
  Surv(time = years_in_trial, event = death) ~ treatment_arm,
  data = cooties
  )
```

```{r}
ggsurvplot(survival,
           linetype = 'strata',
           conf.int = T,
           pval = T)
```

The effect of `sex` within the treatment arm

```{r}
survival <- survfit(
  Surv(time = years_in_trial, event = death) ~ sex,
  data = filter(cooties, treatment_arm == 1)
  )
```

```{r}
ggsurvplot(survival,
           linetype = 'strata',
           conf.int = T,
           pval = T)
```

What other interesting trends exist in the data?

